---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# rmdd

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-green.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![CRAN status](https://www.r-pkg.org/badges/version/rmdd)](https://CRAN.R-project.org/package=rmdd)
[![](http://cranlogs.r-pkg.org/badges/grand-total/rmdd?color=green)](https://cran.r-project.org/package=rmdd)
[![](http://cranlogs.r-pkg.org/badges/last-week/rmdd?color=green)](https://cran.r-project.org/package=rmdd)
<!-- badges: end -->

## Overview

`rmdd` is a tidyverse-style R package that wraps the **Mammal Diversity Database (MDD v2.4)**, the authoritative taxonomic checklist maintained by the American Society of Mammalogists. The package bundles the complete MDD release as ready-to-use tibbles and provides a multi-stage name reconciliation pipeline to resolve mammal scientific names — including synonyms, original combinations, and typographic variants — against the current accepted taxonomy.

The MDD currently covers **6,871 mammal species** (6,758 living wild + 113 recently extinct post-1500 CE), with over **64,683 synonymous names** curated across 27 orders, 167 families, and 1,360 genera.

---

## Installation

```r
# Install from CRAN:
install.packages("rmdd")

# Install the development version from GitHub:
# install.packages("pak")
pak::pak("PaulESantos/rmdd")
```

---

## Bundled datasets

`rmdd` ships three datasets that are loaded lazily and available immediately after `library(rmdd)`:

| Dataset | Rows | Description |
|---|---|---|
| `mdd_checklist` | 6,871 | Species-level checklist with 52 variables (taxonomy, distribution, IUCN status, type locality, cross-release flags) |
| `mdd_synonyms` | ~64,683+ | Nomenclatural synonym table with 44 variables linking every known name combination to its accepted species |
| `mdd_type_specimen_metadata` | 138 | Institutional metadata for natural history collections cited in the synonym table |

All column names are normalized from the original camelCase to `snake_case` on import. The original column names are documented in the help pages for each dataset.

---

## Core functions

### Name reconciliation

```{r}
library(rmdd)

# Reconcile a vector of names against MDD
mdd_matching(c("Puma concolor", "Felis concolor", "Pumma concolor"))

# Retrieve full structured information for a taxon
mdd_taxon_info("Vicugna vicugna")

```

The reconciliation pipeline in `mdd_matching()` follows a waterfall of five stages, moving from strict to relaxed evidence:

1. **Direct match** — exact match on the cleaned full name or genus + epithet pair
2. **Exact genus** — genus found in the MDD, regardless of species
3. **Fuzzy genus** — genus matched by string distance (OSA by default, configurable)
4. **Exact species within matched genus** — species epithet matched exactly within the resolved genus
5. **Fuzzy species within matched genus** — species epithet matched by string distance

Every row in the output carries a `match_stage` label, pathway flags (`direct_match`, `fuzzy_match_genus`, …), fuzzy distance columns, and the full accepted-name context (`accepted_id`, `accepted_name`, `accepted_author`).

### Parsing and classification

```{r}
# Parse a vector of names into taxonomic components
classify_mammal_names(c(
  "Puma concolor",
  "Capromys (Pygmaeocapromys) angelcabrerai",
  "Mus musculus domesticus"
))


mdd_matching("Cricetus cricetus")
```

### Distribution analysis

```{r}
# Country-level mammal diversity summary
mdd_distribution_summary(level = "country")

# Continent-level, excluding domesticated species
mdd_distribution_summary(
  level = "continent",
  exclude_domesticated = TRUE
)

# Distribution map (requires rnaturalearth + rnaturalearthdata)
mdd_distribution_map("Tremarctos ornatus", zoom = "auto")
mdd_distribution_map("Lama vicugna", quiet = TRUE)
```

### Data citations

```{r}
# Formatted MDD dataset citation
mdd_reference()

```

---

## Example workflow

```{r}
library(rmdd)
library(tidyverse)

# 1. A list with spelling variants and synonyms
names_to_check <- c(
  "Panthera onca",
  "Felis onca",          # historical combination
  "Panthera onkca",      # typo
  "Lontra longicaudis",
  "Lutra longicaudis"    # synonym
)

# 2. Run the reconciliation pipeline
result <- mdd_matching(names_to_check, 
                       allow_duplicates = TRUE)

# 3. Inspect results
result |>
  select(input_name, matched_name, taxon_status,
         accepted_name, matched, match_stage)

# 4. Full taxon profile for one species
mdd_taxon_info("Panthera onca")

# 5. Distribution map
mdd_distribution_map("Panthera onca", zoom = "auto")
```

---

## The Mammal Diversity Database

The MDD is maintained by the American Society of Mammalogists and tracks all taxonomic changes to living and recently extinct mammal species (post-1500 CE) in real time. The current release is **v2.4** (January 2, 2026).

Key MDD v2.4 statistics:

| Taxa | Count |
|---|---|
| Total species | 6,871 |
| Living wild species | 6,741 |
| Recently extinct (post-1500 CE) | 113 |
| Domestic species | 17 |
| Genera | 1,360 |
| Families | 167 |
| Orders | 27 |
| Synonymous names | 64,683+ |

The MDD v2 introduced geographic distribution codings at the country, US state, continent, and biogeographic realm levels, a comprehensive nomenclatural synonym dataset with type locality and specimen information curated in coordination with [Hesperomys](https://hesperomys.com) and [Batnames](https://batnames.org), and a redesigned website with advanced search and data visualization.

---

## Citation

### Citing this package

```{r}
citation("rmdd")
```

### Citing the MDD dataset

```
Mammal Diversity Database. (2026). Mammal Diversity Database (Version 2.4) [Data set].
Zenodo. https://doi.org/10.5281/zenodo.17033774
```

Use `mdd_reference()` to generate a correctly formatted citation from within R, including taxon-level entry citations when needed.

### Citing the MDD taxonomy (v2.0)

Burgin, C. J., Zijlstra, J. S., Becker, M. A., Handika, H., Alston, J. M., Widness, J., Liphardt, S., Huckaby, D. G., and Upham, N. S. (2025). How many mammal species are there now? Updates and trends in taxonomic, nomenclatural, and geographic knowledge. *Journal of Mammalogy* in press. <https://doi.org/10.1101/2025.02.27.640393>

### Citing the MDD taxonomy (v1.0)

Burgin, C. J., Colella, J. P., Kahn, P. L., and Upham, N. S. (2018). How many species of mammals are there? *Journal of Mammalogy* 99: 1–11. <https://doi.org/10.1093/jmammal/gyx147>


---

## Related resources

- [MDD website](https://www.mammaldiversity.org/)
- [MDD Zenodo archive](https://doi.org/10.5281/zenodo.4139722)

---

## Contributing and issues

Bug reports and feature requests are welcome at <https://github.com/PaulESantos/rmdd/issues>.
