---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# rmdd

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-green.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![CRAN status](https://www.r-pkg.org/badges/version/rmdd)](https://CRAN.R-project.org/package=rmdd)
[![](http://cranlogs.r-pkg.org/badges/grand-total/rmdd?color=green)](https://cran.r-project.org/package=rmdd)
[![](http://cranlogs.r-pkg.org/badges/last-week/rmdd?color=green)](https://cran.r-project.org/package=rmdd)
[![R-CMD-check](https://github.com/PaulESantos/rmdd/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PaulESantos/rmdd/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/PaulESantos/rmdd/graph/badge.svg)](https://app.codecov.io/gh/PaulESantos/rmdd)
<!-- badges: end -->

`rmdd` provides packaged Mammal Diversity Database (MDD) data for R and a
practical workflow for mammal name reconciliation, accepted-name retrieval,
distribution summaries, and distribution maps. It is designed for analysts who
need a local, reproducible interface to the current MDD release.

## Installation

```r
install.packages("rmdd")
```

Development version:

```r
# install.packages("pak")
pak::pak("PaulESantos/rmdd")
```

## What is included

`rmdd` ships three lazily loaded datasets:

| Dataset | Rows | Purpose |
|---|---:|---|
| `mdd_checklist` | 6,871 | Current accepted species checklist with taxonomy, status, and distribution fields |
| `mdd_synonyms` | 64,683+ | Synonym table linking historical and alternate names to accepted species |
| `mdd_type_specimen_metadata` | 138 | Collection-level metadata for type specimen institutions cited in MDD |

## Quick start

```{r}
library(rmdd)
library(dplyr)

mdd_matching(c("Puma concolor", "Felis concolor", "Pumma concolor")) |>
  select(input_name, matched_name, taxon_status, accepted_name, match_stage)
```

```{r}
mdd_taxon_record("Vicugna vicugna")$taxon_tbl |>
  select(
    sci_name,
    original_name_combination,
    country_distribution,
    iucn_status
  )
```

```{r}
mdd_distribution_summary(level = "country") |>
  arrange(desc(total_species)) |>
  slice_head(n = 5)
```

```{r, fig.alt = "Distribution map of Lama vicugna across western South America."}
mdd_distribution_map("Lama vicugna", quiet = TRUE)
```

## Site structure

The pkgdown site is organized around three entry points:

- **Home**: short package overview, installation, and a minimal workflow.
- **Get started**: a complete walkthrough from bundled datasets to name matching,
  taxon retrieval, and distribution mapping.
- **Reference**: function documentation grouped by data access, reconciliation,
  taxon retrieval, and distribution tasks.

For the full workflow, see the
[Get started article](https://paulesantos.github.io/rmdd/articles/rmdd.html).

## Citation

Use `citation("rmdd")` to cite the package and `mdd_reference()` to cite the
bundled MDD release used by the package.

```{r}
mdd_reference()
```

## Related resources

- [Mammal Diversity Database website](https://www.mammaldiversity.org/)
- [MDD Zenodo archive](https://doi.org/10.5281/zenodo.4139722)

