---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# rmdd

`rmdd` is a development-stage R package for mammal taxonomic name resolution
using the Mammal Diversity Database (MDD).

## Goal

The package is intended to provide a mammal-focused workflow analogous to
plant-oriented TNRS tools, while following tidyverse package conventions and
CRAN-oriented development practices.

## Installation

When `rmdd` is available on CRAN, install it with:

```{r eval = FALSE}
install.packages("rmdd")
```

Install the development version from GitHub with:

```{r eval = FALSE}
# install.packages("pak")
pak::pak("PaulESantos/rmdd")
```

## Planned features

- Download and cache current MDD source files
- Read MDD exports into tidy tibbles
- Standardize input names
- Perform exact and approximate matching
- Report accepted names, unmatched names, and match diagnostics

## Current bootstrap

The project currently includes:

- Core package metadata and license files
- Initial functions for download, load, and simple matching
- Unit tests using `testthat`
- `pkgdown` and `lintr` configuration stubs

## Example

```{r}
library(rmdd)

sample_path <- system.file("extdata", "mdd_sample.csv", package = "rmdd")
mdd_tbl <- mdd_load(sample_path)

mdd_match_names(
  names = c("Puma concolor", "Vicugna vicugna", "Puma concolr"),
  data = mdd_tbl,
  method = "agrep"
)
```

## Reference

```{r}
mdd_reference()
```
